---
title: "Ageing, Dementia and TBI Data Analysis"
author: "ZHAO Kai"
fontsize: 10pt
output: pdf_document
---

Here is a brief introduction of analyzing the results of our proposed approach on [Ageing, Dementia and TBI dataset](https://aging.brain-map.org).

##  Preparations
```{r, size="small"}
suppressPackageStartupMessages({
  require(ggplot2)
  require(formatR)
  require(knitr)
  require(cluster)
  require(factoextra)
  require(dplyr)
  require(RColorBrewer)
  require(clusterProfiler)
  require(org.Hs.eg.db)
  require(enrichplot)
  require(stringr)
  require(forcats)
  require(ggplot2)
  require(ggdendro)
  require(graphics)
  require(gridExtra)
  require(extrafont)
  require(viridis)
  require(hrbrthemes)
})

truncated_var <- function(x){
    remove_idx <- c(which.max(x), which.min(x))
    var(x[-remove_idx])
}

wrap_labal <- function(x, width = 60){
    str_wrap(x, width=60)
}

simes.test <- function(x, returnstat = FALSE){ 
    r = rank(x,  ties.method = "random") 
    t = min(length(x) * x / r) 
    if (returnstat) c(t, t) else t 
} 
```

```{r}
setwd("/Users/zhaokai/data/Results/ageing")

load("ageing_dataset_annotated_with_phenotypes_filtered.RData")
pheno <- dataset[,2]
tissue <- dataset[,3]
dataset <- dataset[, -(1:4)]
# our fitted model 
load("insider_ageing_R23_fitted_object.RData")
# load("gene_expression_iMF_L1_penalty_25v2.RData")
# attach(fitted_obj)
attach(object) 
disease_factor <- cfd_matrices[[1]]
tissue_factor <- cfd_matrices[[2]]
donor_factor <- cfd_matrices[[3]]
interactions <- cfd_matrices[[4]]

# read meta information to facilitate our analyis
meta <- read.csv("meta_info.csv", header = TRUE, stringsAsFactors = F)
gene_info <- read.csv("rows-genes.csv", header = TRUE, stringsAsFactors = F)
structure_info <- unique(read.csv("structure_id_mapping.csv", stringsAsFactors = F)[, c(14, 15)])

structure_info$snames[c(4,7)] <- c("hippocampus_right", "hippocampus_left")

# match gene_info with genes included in the study
gene_included <- data.frame(gene_id = as.numeric(gsub("X", "", colnames(data))))
gene_info_inc <- inner_join(gene_included, gene_info, by = "gene_id")

row.names(tissue_factor) <- structure_info$snames
row.names(disease_factor) <- c("ctrl", "case")
# head to get a sense of our results and meta information
# str(fitted_obj)
head(meta)
head(gene_info)
```

##  Cluster analysis of donor_factor

The donor_factor is a low rank representation of genetic information from the dementia dataset, which is a matrix of $N$ rows and $K$ columns. $N$ is the number of donors and $K$ is the number of latent features representing gene expression information in low dimensions. In this section, the cluster analysis basically follows this [tutorial](uc-r.github.io/hc_clustering).

We exclude a number of metagenes that is irrelevant to synapse function, cognition, and memory. The selection of metagenes is somewhat subjective. However, we observe that this strategy shows some interesting results. 

```{r, fig.height=6, size="small"}
# for detail, see https://uc-r.github.io/hc_clustering methods to assess
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(donor_factor, method = x)$ac
}
ac_vec <- sapply(m, function(x) ac(x))
ac_vec

# carry out hierarchical cluster analysis using the method with the greatest coefficient
hc3 <- agnes(donor_factor[,-c(14)], method = unname(m[which(ac_vec == max(ac_vec))]))
hc <- as.hclust(hc3)
ggdendrogram(hc, theme_dendro = FALSE) +
  ggtitle("Dendrogram of donor clustering") + ylab("Height") +
  theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
  theme(plot.title = element_text(hjust = 0.5, size=16,face="bold"), 
        text=element_text(family="Times New Roman"), 
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# select cluster number using Gap statistics
# gap_stat <- clusGap(donor_factor, FUN = hcut, nstart = 25, K.max = 10, B = 200)
# fviz_gap_stat(gap_stat)

# choose 3 clusters by investigating Dendrogram and Gap statistics plots 
sub_grp <- cutree(hc3, k = 2)
result <- cbind(meta, cluster_id = sub_grp)
aov_res <- aov(age ~ cluster_id, data = result)
unname(unlist(summary(aov_res)))[9]

kruskal.test(age ~ cluster_id, data = result)
fisher.test(table(result$cluster_id, result$braak), simulate.p.value=TRUE)
```


##  explore the relavance of the clustering

In this section, we only examined the association between the age of donors and their clusters. Explorations of the relavance of the clustering to other clinical variables can also be carried out. Pie charts and histograms can be drawn to visualize the association. Furthermore, some statistical tests can also be used to check the significance. The result below shows that there is a statistical significant association between the age and clustering. 

```{r}
exc_idx <- c(4, 5, 8, 19)
# carry out hierarchical cluster analysis using the method with the greatest coefficient
hc3 <- agnes(donor_factor[,-exc_idx], method = unname(m[which(ac_vec == max(ac_vec))]))
hc <- as.hclust(hc3)
ggdendrogram(hc, theme_dendro = FALSE) +
  ggtitle("Dendrogram of donor clustering") + ylab("Height") +
  theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
  theme(plot.title = element_text(hjust = 0.5, size=16,face="bold"), 
        text=element_text(family="Times New Roman"), 
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# choose 3 clusters by investigating Dendrogram and Gap statistics plots 
sub_grp <- cutree(hc3, k = 3)
result <- cbind(meta, cluster_id = sub_grp)
aov_res <- aov(age ~ cluster_id, data = result)
unname(unlist(summary(aov_res)))[9]

kruskal.test(age ~ cluster_id, data = result)
fisher.test(table(result$cluster_id, result$braak), simulate.p.value=TRUE)

fisher.test(table(result$cluster_id, result$nincds_arda_diagnosis)[,-1])
fisher.test(table(result$cluster_id, result$dsm_iv_clinical_diagnosis)[,-c(4,5,6)])
```

The below histogram show the dribution of clusters cross different braak stages, which are clinical diagnoses of stage of dementia.

```{r, fig.height=6, size="small"}
cluster_tbi <- table(result$cluster_id, result$braak)
par(mar=c(5, 13, 3, 13))
barplot(cluster_tbi, main="number of observations",
  xlab="braak", col= brewer.pal(3, "Set2") ,
  legend = rownames(cluster_tbi), space = 0.2, width = 0.2, 
  args.legend = list(x = "toprigh"))
```

## Enrichment analysis of biological processes involved
We can also investigate biological processes we are interested in with our results. For example, we explored the mechanism of dementia with the disease and gene factors. We obtained the expression profiles for the dementia and control, extracted the genes with greatest postive difference between them, and examine the biological processes enriched by those genes. 

```{r, fig.height=6, size="small"}
idx <- order(apply(disease_factor, 2, function(x) abs(x[1] - x[2])), decreasing = T)[1:2] 
disease_matrix <- disease_factor[,idx, drop = F] %*% column_factor[idx, , drop = F]
diff <- disease_matrix[2,] - disease_matrix[1,]

cutoffs <- quantile(diff, probs = seq(0, 1, 0.025))
selected <- (diff <= cutoffs[2])  
# selected <- (column_factor[idx,] >= cutoffs[length(cutoffs)-1]) # greatest negative difference
# cutoffs <- quantile(column_factor[idx,], probs = seq(0, 1, 0.025))
# selected <- (column_factor[idx,] <= cutoffs[2]) # greatest negative difference

down_reg <- enrichGO(gene      = unique(gene_info_inc[selected,3]),
                     OrgDb     = 'org.Hs.eg.db',
                     ont       = "BP",
                     readable  = TRUE)

dotplot(down_reg, font.size = 8, showCategory=30, label_format = 80) +
  theme(text=element_text(family="Times New Roman"))
```

## compare expression level across different brain structures for selected genes 

```{r, fig.height=12, fig.width = 12, size="small"}
loadfonts(quiet = T)

structure_names <- c("TC left", "FWM right", "FWM left", "HPC right", "PC right", "TC right", "HPC left", "PC left")

# result <- as.data.frame(upreg)
# gene_names <- unlist(strsplit(result[1,8], split = "/"))

selected <- c("PLEKHG5", "NCS1", "GRIK5", "NRGN")

row_ids <- unlist(sapply(selected, function(x) which(gene_info_inc[[4]] == x)))
data <- cbind(pheno, structure_names[tissue], dataset[,row_ids])
colnames(data) <- c("pheno", "tissue", names(row_ids))
data$pheno <- as.factor(data$pheno)

# boxplots and t-tests for the 4 variables at once
test_results <- sapply(3:ncol(data), function(j){
    pvalues <- sapply(structure_names, function(i) t.test(data[data$tissue == i, j] ~ data$pheno[data$tissue == i])$p.value)
})

simes_pvalues <- apply(test_results, 2, function(x)simes.test(x))
names(simes_pvalues) <- names(row_ids)

colnames(test_results) <- names(row_ids)
rownames(test_results) <- structure_names

p1 <- ggplot(data, aes(x=tissue, y=PLEKHG5, fill=pheno)) + 
    scale_color_viridis(discrete = TRUE) +
    geom_boxplot() + 
    ggtitle("PLEKHG5 expression") +
    xlab("") + ylab("Levels") +
    theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

p2 <- ggplot(data, aes(x=tissue, y=NCS1, fill=pheno)) + 
    scale_color_viridis(discrete = TRUE) +
    geom_boxplot() + 
    ggtitle("NCS1 expression") +
    xlab("") + ylab("Levels") +
    theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

p3 <- ggplot(data, aes(x=tissue, y=GRIK5, fill=pheno)) +
    scale_color_viridis(discrete = TRUE) +
    geom_boxplot() + 
    ggtitle("GRIK5 expression") +
    xlab("") + ylab("Levels") +
    theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

p4 <- ggplot(data, aes(x=tissue, y=NRGN, fill=pheno)) +
    scale_color_viridis(discrete = TRUE) +
    geom_boxplot() + 
    ggtitle("NRGN expression") +
    xlab("") + ylab("Levels") +
    theme_ipsum(base_family = "Times New Roman", base_size= 12, plot_title_face = "bold", axis_title_size = 16, plot_title_size = 18) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

grid.arrange(p1, p2, p3, p4, nrow=2, ncol=2) 
```

Here is a another pieces of code to demenstrate molecular functions of genes with the largest effects in different brain structures. 

\textcolor{red}{In this part of analysis, I only demonstrate with the second tissue. In order to expand, analysis of other metagenes with a single for loop is fine to generate all results.}

```{r, fig.height=6, size="small"}
idx <- order(apply(tissue_factor, 2, function(x) truncated_var(x)),decreasing=T)[1:3]
tissue_matrix <- tissue_factor[,idx] %*% column_factor[idx, ]
row.names(tissue_matrix) <- structure_info$snames
id <- 2 
cat("tissue name:", row.names(tissue_matrix)[id], "\n")
cutoffs <- quantile(tissue_matrix[id, ], probs = seq(0, 1, 0.025)) 
selected <- (tissue_matrix[id, ] >= cutoffs[length(cutoffs)-1])  # greatest positive difference

# head(gene_info[selected,3])
up_reg <- enrichGO(gene      = unique(gene_info_inc[selected,3]),
                   OrgDb     = 'org.Hs.eg.db',
                   ont       = "BP",
                   readable  = TRUE)

dotplot(up_reg, font.size = 9, showCategory=30, label_format = 50) +
  theme(text=element_text(family="Times New Roman"))

selected <- (tissue_matrix[id,] <= cutoffs[2]) # greatest negative difference
down_reg <- enrichGO(gene    = unique(gene_info_inc[selected,3]),
                   OrgDb     = 'org.Hs.eg.db',
                   ont       = "BP",
                   readable  = TRUE)

dotplot(down_reg, font.size = 9, showCategory=30, label_format = 60) +
  theme(text=element_text(family="Times New Roman"))
```

Furthermore, we could also examine the interaction between tissue and disease factors. The code below explores the interaction between different tissues and dementia. Then, similar techniques can be employed to examine the contribution of underlying biological processes to the interaction. 

```{r, fig.height=6, size="small"}
# since all latent vectors are restricted in the same space, we can compute the correlation between disease and tissue factos. 
confounder <- object[['confounder']][,-3]
unique_cfd <- unique(confounder)

distances <- sapply(seq(8), function(i){
  idx <- which(unique_cfd[,2] == i)
  # mean((interactions[idx,1] - interactions[idx,2])^2)
  t.test(interactions[idx,1], interactions[idx,2], paired = T)$p.value
})

# then we can examine the tissue with the largest distance between dementia and control
tissue_id <- which.min(distances)  
cat("Tissue name:", structure_info$snames[tissue_id], "\n")

cfd_id <- which(unique_cfd[,2] == tissue_id)
```

```{r, fig.height=6, size="small"}
selected <- order(apply(interactions[cfd_id, ], 2, var), decreasing = TRUE)[1:3]

interaction_matrix <- interactions[cfd_id, selected, drop = F] %*% column_factor[selected, , drop = F]
diff <- interaction_matrix[2, ] - interaction_matrix[1,]

cutoffs <- quantile(diff, probs = seq(0, 1, 0.025))

# up-regulation, greatest positive difference
selected <- (diff >= cutoffs[length(cutoffs)-1])  
upreg <- enrichGO(gene      = unique(gene_info_inc[selected,3]),
                  OrgDb     = 'org.Hs.eg.db',
                  ont       = "BP",
                  readable  = TRUE)

dotplot(upreg, font.size = 8, showCategory=30, label_format = 70) + 
  theme(text=element_text(family="Times New Roman"))

# down-regulation, greatest negative difference
selected <- (diff <= cutoffs[2])
downreg <- enrichGO(gene    = unique(gene_info_inc[selected,3]),
                    OrgDb     = 'org.Hs.eg.db',
                    ont       = "BP",
                    readable  = TRUE)

dotplot(downreg, font.size = 9, showCategory=30, label_format = 50) + 
  theme(text=element_text(family="Times New Roman"))
```